Linear and Differential Cryptanalysis of DES

### 2009-01-29 trunk/talk/draft1.tex

\begin{frame}
How to exploit those - Linear attack idea

@@

@@

@@

\subsection{Linear Attack on SPN}
\begin{frame}
Basic math concepts behind Linear attack ...

>>>>

\begin{frame}
Linear Attack on DES idea:\\
\begin{itemize}
\item S-Boxes depend on a relatively small number of bits (6) which allows us to write down linear (or afine) expressions that approximate S-Boxes \\
\item The effects of one round do not diffuse quickly over following rounds. Thus linear expresions (as above) that hold per-round can be combined across rounds.
\end{itemize}
\end{frame}

\begin{frame}
Specifically, if $P_{i}$ are plaintext bits, $C_{i}$ are cyphertext bits, and $K_{i}$ are subkey bits, then we wish to find an expression of the form
\[ P_{i_1} \oplus P_{i_2} \cdots P_{i_j} \oplus C_{i_1} \oplus C_{i_2} \cdots C_{i_k} = K_{i_1} \oplus K_{i_2} \oplus K_{i_m} \]
Such that this expression has a high \textit{or} low probability of occurence.\\
\vspace{5mm}
Consider: No such obvious expression should exist, otherwise the cipher is trivially weak. If we were to randomly select bits for the above expression, it would hold exactly $1/2$ the time. If we find an expression such as above that displays a high \textit{bias}, that is, it holds much more or less freqently than $1/2$ the time, we can exploit this.
\end{frame}

@@

We can remove much of the complexity of DES without affecting the discussion of the techniques of Linear Cryptanalysis. We define a simple Substitution-Permutation Network Cipher (SPN) which has structural similarity to DES.

\begin{itemize}
\item SPN is a 16-bit block cipher (16-bit plaintext to 16-bit ciphertext) with 16-bit round keys.
\item Each round consists of a substitution and a permutation (much like in DES)
\item The substitution is the result of splitting the 16-bit input block to a round in to 4 4-bit sub-blocks. Each 4-bit sub-block is then mapped across a 4-bit to 4-bit S-box. We use the same S-box throughout. 
\item The permutation is applied to all 16-bits of the round following the substitution.
\item Finally, key-mixing is achieved by XOR-ing the round key with every input block to a round as well as at the end of the last round (so that we can't simply ignore the last round)
\item The SPN cipher we will look at will consist of 4 rounds.
\end{itemize}

@@

\begin{frame}
SPN Visually
%\pgfimage[scale=0.75]{SPN.PNG}
\end{frame}

@@

\subsection{Linear Attack on SPN}
\begin{frame}
\large{Basic math concepts behind the Linear attack:}\\
\vspace{5 mm}
Linear cryptanalysis tries to take advantage of high probability occurrences of  linear expressions involving plaintext bits, ciphertext bits and subkey bits.  It is a known plaintext attack.\\
\vspace{5 mm}
Define $P_i$ where $i = 1, 2, \dots 16$ as the $i$-th plaintext bit.\\
Define $C_i$ where $i = 1, 2, \dots 16$ as the $i$-th cyphertext bit.\\
Define $U_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th input bit to the $j$-th round of SPN.\\
Define $V_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th output bit of the $j$-th round of SPN.\\
\end{frame}
